---
title: "Intoduction to DomArchive"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How-to}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`DomArchive` is a data package compiling dominance interaction data published
over a century of research. The goal of the package is to facilitate comparative
research on the structure and function of dominance hierarchies. This vignette
provides guidance on how to interact with the data objects in the R package. 

## DomArchive data objects

`DomArchive` contains the following data objects: 

1. `dom.data` - a list containing both data and metadata. Each element of the list
is a unique data entry, named after the first author and year of publication.
Each element of the list contains two named elements -- `matrix` containing the 
interaction sociomatrix and `metadata` containing the metadata associated with the 
data. 

2. `dom.metadata` - a dataframe containing the metadata for the complete dataset. 
The information in the `dom.metadata` dataframe is exactly identical to what you would get if you
extracted and combined every `metadata` element in `dom.data`. 

As an example, here is how to extract the sociomatrix and metadata associated with Schjelderup-Ebbe's 
1922 paper: 

```{r, results='hide'}
library(DomArchive)
data(dom.data)
data(dom.metadata)
## Extract sociomatrix
dom.data$SchjelderupEbbe_1922a$matrix

## Extract metadata from `dom.data`
dom.data$SchjelderupEbbe_1922a$metadata

## Extract metadata from `dom.metadata`
dom.metadata[dom.metadata$fileid == 'SchjelderupEbbe_1922a',]
```

Here is a summary of the metadata associated with each datafile:

|Metadata Column| Meaning                           | Potential values
|---------------|-----------------------------------|-----------------
|fileid         | Unique identifier for data        |
|taxon          | Order                             |
|species        | Species name                      |
|common_name    | Common name                       |
|study_site     | Nation where study was conducted  |
|captivity      | Captive or free-ranging animals?  | "Captive", "Natural"
|sex            | Males, females, or both?          | "M", "F", "MF"
|age            | What age classes                  | "Adult", "Non-Adult", "Mixed"
|measure        | What behavior was measured?       |
|matrix_location|                                   |
|countbinary    | Are data raw counts or binary?    | "count", "binary"
|groupid        | Unique identifier for social group|

Additionally, metadata includes some quantities calculated from the interaction data. 
Calculations were made using functions in the `compete` package (https://github.com/jalapic/compete). 

|Quantity                    | Details                           
|----------------------------|------------------------------------
|number_individuals          | Number of individuals in the data
|number_interactions         | Number of interactions in the data
|interactions_per_individual | Number of interactions per individual
|proportion_unknown          | Proportion of relationships for which there are no observations
|dci                         | Directional consistency index (Van Hooff & Wensing 1987)
|ttri                        | Triangle transitivity index (Shizuka & McDonald 2012)
|ds_steepness                | Hierarchy steepness calculated from David's Scores (de Vries et al 2006)
|modified_landaus_h          | Modified Landau's h` measure of linearity (de Vries 1995)

## Subsetting the archive using `subset_archive()`

`DomArchive` includes one function for interacting with the data objects. The 
`subset_archive()` function allows users to extract specific data elements from
`dom.data`. There are two ways users can specify which data they would like to retain. 

1. Users can provide a vector of fileids. `subset_archive()` will return data associated with
those fileids. 

2. Users can provide a list of column names and a list of values to be matched in those columns. 
`subset_archive()` will return all rows where the column includes the provided values. To specify multiple
restriction criteria, users can provide`column` and `value` as lists, and the
each column is matched to the corresponding value(s) (i.e., check that for each row `column[n] %in% values[n]`).

If both `fileids` and `column`/`values` are supplied, only `fileids` will be used. 
Additionally, the `return.intxdata.only` flag can be used to make `subset_archive()`
return only interaction data instead of interaction data and metadata. If users
would like to do this on their own, this flag is the equivalent of using `map` from
`purrr` to select only the data matrices: 

`purrr::map(dom.data, 'matrix')`

For more details, see Example analysis or ?subset_archive(). 

## Example analysis

Here we are interested in how directional consistency differs between natural primate
groups and wild primate groups. Does housing primates in captivity influence the 
structure of social hierarchies? We have an intuition that directional consistency
might be influenced by group size, so we will want to control for group size in our 
analysis. For this example we will use the matrix data directly, although 
this information is already contained in `dom.metadata`. 

First, we will restrict the dataset to only primates using the `taxon` column. Here we
do this with `subset_archive()` using both approaches described above. 

```{r}
## Option 1 - using fileids
captive.fileids <- dplyr::filter(dom.metadata, taxon == 'Primate', captivity == 'Captive')$fileid
primates.captive <- subset_archive(fileids = captive.fileids, return.intxdata.only = TRUE)
natural.fileids <- dplyr::filter(dom.metadata, taxon == 'Primate', captivity == 'Natural')$fileid
primates.natural <- subset_archive(fileids = natural.fileids, return.intxdata.only = TRUE)


## Option 2 - using 'column' and 'values'
primates.captive <- subset_archive(column = list('taxon', 'captivity'),
                                   values = list('Primate', 'Captive'),
                                   return.intxdata.only = T)

primates.natural <- subset_archive(column = list('taxon', 'captivity'),
                                   values = list('Primate', 'Natural'),
                                   return.intxdata.only = T)


```

Now that we have extracted matrices for primates in captivity and in the wild, we 
are ready to extract some features from the matrices. Here we extract the group size
and the directional consistency index.

```{r}
captivity <- data.frame(dci = purrr::map_dbl(primates.captive, compete::dci),
                        num.ids = purrr::map_dbl(primates.captive, nrow))

natural <- data.frame(dci = purrr::map_dbl(primates.natural, compete::dci),
                        num.ids = purrr::map_dbl(primates.natural, nrow))

```

Finally, let's plot the relationship between these variables. 

```{r, fig.align='center', fig.height=4, fig.width= 6, echo = FALSE}
mfrow = c(1,1)
plot(captivity$num.ids, captivity$dci, col = 'darkorchid4', pch = 16, 
     xlab = 'Number of individuals', ylab = 'Directional Consistency Index')
captivity.coefs <- lm(data = captivity, dci ~ num.ids)$coefficients
abline(a = captivity.coefs[1], b = captivity.coefs[2], col = 'darkorchid4', lwd = 2)

points(natural$num.ids, natural$dci, col = 'firebrick', pch = 16)
natural.coefs <- lm(data = natural, dci ~ num.ids)$coefficients
abline(a = natural.coefs[1], b = natural.coefs[2], col = 'firebrick', lwd = 2)
legend(pch = 16, legend = c('Captive', 'Natural'),
       col = c('darkorchid4', 'firebrick'), x = 'bottomright')

```


## References
1. Schjelderup-Ebbe, T. (1922). Beitrage zur sozialpsychologie des haushuhns. Zeitschrift Fur Psychologie Und Physiologie Der Sinnesorgane, 88, 225–252.

2. Curley, J. compete: Analyzing Social Hierarchies. R package 
version 0.1. https://github.com/jalapic/compete

3. Van Hooff, J., Wensing J.A.B. (1987). Dominance and its behavioural measures in a captive wolf pack. In: Frank H.W., editor. Man and Wolf. Dordrecht, Olanda (Netherlands): Junk Publishers pp.219-252.

4. , D., & McDonald, D. B. (2012). A social network perspective on measurements of dominance hierarchies. Animal Behaviour, 83(4), 925–934. http://dx.doi.org/10.1016/j.anbehav.2012.01.011

5. de Vries, H., Stevens, J. M. G., & Vervaecke, H. (2006). Measuring and testing the steepness of dominance hierarchies. Animal Behaviour, 71(3), 585–592. http://webhost.ua.ac.be/crc/pdfs/DeVriesAB2006.pdf

6. de Vries, H. (1995). An improved test of linearity in dominance hierarchies containing unknown or tied relationships. Animal Behaviour, 50(5), 1375–1389. http://www.sciencedirect.com/science/article/pii/0003347295800530